---
title: "R-Packages"
author: "Amitai Asaf"
date: "2023-12-26"
output: html_document
---

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load librarys and data

```{r librarys, results="hide", error=FALSE,warning=FALSE,message=FALSE}
library(tidytuesdayR)
library(tidyverse)
library(igraph)
library(ggplot2)

last_tuesday()
r_data <- tt_load("2023-12-26")
```

# inspect data

```{r glimpse}
map(r_data,glimpse)
```

# SNA

### Focus

The three data sets provide a lot of data as expected from a database collecting data about R packages. This could be a bit overwhelming, but with a small glimpse the second data frame catches the eye. The second data frame, also called external calls, lists the amount of calls each R package makes to another R-package.

This type of data could also be thought as an edge list, a list of connections between nodes. Nodes are actors who exist either in the world or as a concept, and they could have relations with other nodes. Together the actors (nodes) and the connections between them (edges) can form a network of connections also known as a social network. 

Here I am going to analyze the social network of R-packages and try to learn about the relationships between the packages.

## create basic social network

First we create the basic network

```{r create network}

external_calls <- r_data$external_calls

sn <- graph_from_edgelist(cbind(external_calls$package_from,external_calls$package_to))
E(sn)$weight <- external_calls$n_total
```

Then we check how many actors and connections are there in the network
```{r asses network}
paste("There are",as.character(vcount(sn)) , "actors")
paste("and" , ecount(sn) , "connections between them in the network")
```
This makes the network too big for a plot but we can still analyze it.

## centralization measurments

```{r centrality}

sn_centrality <- data.frame(Package = V(sn)$name,
                            in_degree_cen = degree(sn,mode="in"),
                            out_degree_cen = degree(sn,mode="out"),
                            total = degree(sn,mode="total"),
                            in_strength_cen = strength(sn,mode="in"),
                            out_strength_cen = strength(sn,mode="out"),
                            betweeness_cen = betweenness(sn),
                            in_closeness_cen = closeness(sn,mode="in"),
                            out_closeness_cen = closeness(sn,mode="out"),
                            eigen_cen = eigen_centrality(sn,directed = TRUE)$vector)

sn_centrality <-replace(sn_centrality,is.na(sn_centrality),0)
```

### degree

The degree of an actor is simply the amount of connections the actor have to its neighbors.
In degrees represent the amount of packages that calls to a package and out degrees represent the amount of packages a package calls to. 

A weighted degree, also called strength is the amount of calls a package make or receive respectively to in and out degrees.

```{r degree,warning=FALSE}

ggplot(sn_centrality, aes(x = in_degree_cen))+
  geom_bar() +
  xlim(0,50)+
  ylim(0,500)

ggplot(sn_centrality, aes(x = out_degree_cen))+
  geom_bar() +
  xlim(0,500) +
  ylim(0,500)

ggplot(sn_centrality, aes(x = in_strength_cen))+
  geom_bar() +
  xlim(0,500)+
  ylim(0,500)

ggplot(sn_centrality, aes(x = out_strength_cen))+
  geom_bar() +
  xlim(0,500) +
  ylim(0,500)


```

Lets look which packages have the highest degrees and strengths

```{r top degree,warning=FALSE}

sn_centrality %>%
  mutate(Package = fct_reorder(Package, in_degree_cen)) %>%
    arrange(desc(in_degree_cen)) %>%
      head(10) %>%
        ggplot(aes(x=Package, y=in_degree_cen,fill=Package)) +
          geom_col() +
          ylab("In-Degree") +
          coord_flip()

sn_centrality %>%
  mutate(Package = fct_reorder(Package, out_degree_cen)) %>%
    arrange(desc(out_degree_cen)) %>%
      head(10) %>%
        ggplot(aes(x=Package, y=out_degree_cen,fill=Package)) +
          geom_col() +
          ylab("Out-Degree") +
          coord_flip()


sn_centrality %>%
  mutate(Package = fct_reorder(Package, in_strength_cen)) %>%
    arrange(desc(in_strength_cen)) %>%
      head(10) %>%
        ggplot(aes(x=Package, y=in_strength_cen,fill=Package)) +
          geom_col() +
          ylab("In-Strength") +
          coord_flip()


sn_centrality %>%
  mutate(Package = fct_reorder(Package, out_strength_cen)) %>%
    arrange(desc(out_strength_cen)) %>%
      head(10) %>%
        ggplot(aes(x=Package, y=out_strength_cen,fill=Package)) +
          geom_col() +
          ylab("Out-Strength") +
          coord_flip()

```

### Betweeness

Betweenes is a measurement of how many cases an actor stand in the path of 2 other actors.
In our case a package may not call a package directly, rather it could call another package that would call the package instead and thus use it indirectly.

Lets view the packages to stand between other packages the most.

```{r betweeness}
sn_centrality %>%
  mutate(Package = fct_reorder(Package, betweeness_cen)) %>%
    arrange(desc(betweeness_cen)) %>%
      head(10) %>%
        ggplot(aes(x=Package, y=betweeness_cen,fill=Package)) +
          geom_col() +
          ylab("betweeness") +
          coord_flip()
```

### closeness

Closeness measure the sum of shortest distances between one actor and the other actors in means of indirect (or direct) relations

```{r closeness}
sn_centrality %>%
  mutate(Package = fct_reorder(Package, in_closeness_cen)) %>%
    arrange(desc(in_closeness_cen)) %>%
      head(10) %>%
        ggplot(aes(x=Package, y=in_closeness_cen,fill=Package)) +
          geom_col() +
          ylab("in-closeness") +
          coord_flip()

sn_centrality %>%
  mutate(Package = fct_reorder(Package, out_closeness_cen)) %>%
    arrange(desc(out_closeness_cen)) %>%
      head(10) %>%
        ggplot(aes(x=Package, y=out_closeness_cen,fill=Package)) +
          geom_col() +
          ylab("out-closeness") +
          coord_flip()
```

### Eigen Vector

The last measurement is the Eigen Vector. Unlike other measurements the eigen vector is a measurement of the neighboors of an actor rather than the actor itself. It assumes that if an actor is related to a lot of other important actors than it is important itself

```{r Eigen Vector}
sn_centrality %>%
  mutate(Package = fct_reorder(Package, eigen_cen)) %>%
    arrange(desc(eigen_cen)) %>%
      head(10) %>%
        ggplot(aes(x=Package, y=eigen_cen,fill=Package)) +
          geom_col() +
          ylab("Eigen Vector Score") +
          coord_flip()

```

## Plotting a social network

```{r ego}

tidyverse_sn <- induced_subgraph(sn,unlist(ego(sn,nodes = "tidyverse")))


plot.igraph(tidyverse_sn,
            layout= layout_with_dh(tidyverse_sn),
            vertex.size = 20,
            vertex.color = "red",
            vertex.label.cex = .5,
            vertex.frame.color = NA,
            layout = layout.fruchterman.reingold,
            main = "tidyverse ego network"
            )
```